---
title: "erma: epigenomics road map adventures"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "February 2015"
output:
  BiocStyle::pdf_document:
    toc: yes
    number_sections: yes
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{xyz}
-->

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```
```{r dosetup,echo=FALSE,results="hide"}
suppressPackageStartupMessages({
library(BiocStyle)
library(erma)
library(GenomicFiles)
library(ggplot2)
library(grid)
library(png)
})
```

# Introduction

The epigenomics road map describes locations of epigenetic
marks in DNA from a variety of cell types.  Of interest are
locations of
histone modifications, sites of DNA methylation,
and regions of accessible chromatin.

This package presents a selection of elements of the road 
map including metadata and outputs of the ChromImpute
procedure applied to ENCODE cell lines by [Ernst and Kellis](http://dx.doi.org/10.1038/nbt.3157).

# Metadata on the ChromImpute archive

## Sample metadata

I have retrieved a [Google Docs spreadsheet](https://docs.google.com/spreadsheet/ccc?key=0Am6FxqAtrFDwdHU1UC13ZUxKYy1XVEJPUzV6MEtQOXc&usp=sharing#gid=15) 
with comprehensive information.  The mapmeta() function provides
access to a local DataFrame image of the file as retrieved in mid April
2015.

```{r lkm}
library(erma)
meta = mapmeta()
meta
```

We can see the range of tissue types studied:
```{r lktiss}
meta[order(meta[,15]),c(2,15)][1:3,]
sort(meta[,15])
```

## Metadata on the inferred states

The chromatin states and standard colorings
used are enumerated in `states_25`:
```{r lkst}
data(states_25)
head(states_25)
```

The emission parameters of the 25 state model are depicted
in the supplementary Figure 33 of Ernst and Kellis:

```{r dofig,fig=TRUE}
library(png)
im = readPNG(system.file("pngs/emparms.png", package="erma"))
grid.raster(im)
``` 

# Managing access to imputed chromatin states for a set of cell types

I have retrieved a modest number of roadmap bed files with ChromImpute
mnemonic labeling of chromatin by states.  These can be
managed with an ErmaSet instance,
a trivial extension of `r Biocpkg("GenomicFiles")` class.

```{r getgf}
ermaset = makeErmaSet()
ermaset
cellTypes(ermaset)
```

# Enumerating states in the vicinity of a gene, across cell types

We form a GRanges representing 50kb upstream of IL33.
```{r getgm, cache=TRUE}
uil33 = flank(resize(range(genemodel("IL33")), 1), width=50000)
uil33
```
Bind this to the ErmaSet instance.
```{r bind}
rowRanges(ermaset) = uil33
ermaset
```

Now query the files for cell-specific states in this interval.

```{r getcss}
csstates = lapply(reduceByFile(ermaset, MAP=function(range, file) {
  imp = import(file, which=range)
  imp$rgb = rgbByState(imp$name)
  imp
}), "[[", 1) 
tys = cellTypes(ermaset)  # need to label with cell types
csstates = lapply(1:length(csstates), function(x) {
   csstates[[x]]$celltype = tys[x]
   csstates[[x]]
   })
```
Finally, a rough visualization with ggplot2.
```{r doviz, fig=TRUE}
stateProfile(ermaset, symbol="CD28")
```



